Magnetized Polish doughnuts revisited

### 2017-01-17 paper.tex
Some more work on the draft

The investigation of such systems, either by analytical or numerical means, may rely on the ability to construct suitable 
representations based on physical assumptions. The construction of equilibrium models of stationary disks around black holes has a long tradition (see~\cite{Abramowicz:2013} and references therein). In particular, the so-called ``Polish doughnuts", developed by Paczy\'nski and collaborators~\citep{Abramowicz:1978}, provides a  very general method to build equilibrium configurations of perfect fluid matter  
 orbiting around a Kerr black hole. This fully relativistic model assumes that the disk is non-magnetized and that the matter obeys a  constant specific angular momentum distribution. \cite{Komissarov:2006} extended this method by adding a purely azimuthal magnetic 
 field to build magnetized tori around rotating black holes. Assuming different distributions of angular momentum in the disks, \citet{Qian:2009} presented a method to build sequences of black hole thick accretion disks in dynamical equilibria, restricted to the purely hydrodynamical case. Building on this work, we discuss here the extension of the
models of~\citet{Qian:2009} to disks endowed with a purely toroidal magnetic field. For the particular case of constant angular momentum distributions our method shows good agreement with the results of~\citet{Komissarov:2006}.

@@

The equations of ideal general relativistic MHD are $\nabla_{\mu} T^{\mu\nu} = 0$, $\nabla_{\mu} \,^\ast F^{\mu\nu} = 0$, and 
$\nabla_{\mu} (\rho u^{\mu}) = 0$, 
%\begin{equation}\label{eq:e-m_cons}
%\nabla_{\mu} T^{\mu\nu} = 0,
%\end{equation}
%\begin{equation}\label{eq:cons_faraday}
%\nabla_{\mu} \ast F^{\mu\nu} = 0,
%\end{equation}
%\begin{equation}\label{eq:continuity}
%\nabla_{\mu} (\rho u^{\mu}) = 0,
%\end{equation}
where $\nabla_{\mu}$ is the covariant derivative and
\begin{equation}\label{eq:e-m_tensor}
T^{\mu\nu} = (w + b^2)u^{\mu}u^{\nu} + \left(p + \frac{1}{2}b^2\right)g^{\mu\nu} - b^{\mu}b^{\nu},
\end{equation}
is the energy-momentum tensor of a magnetized perfect fluid, $w$ and $p$ being the fluid enthalpy and pressure, respectively. 
Moreover,
%\begin{equation}
$^\ast F^{\mu\nu} = b^{\mu}u^{\nu} - b^{\nu}u^{\mu}$
%\end{equation}
is the (dual of the) Faraday tensor relative to an observer with four-velocity $u^{\mu}$, and $b^{\mu}$ is the magnetic field in that frame.
Assuming the magnetic field is purely azimuthal, i.e.~$b^r = b^{\theta} = 0$,
and taking into account that the flow is stationary and axisymmetric, the conservation of the current density and of the Faraday tensor follow. Contracting Eq.~\eqref{eq:e-m_tensor} with the projection tensor $h^{\alpha}_{\beta} = \delta^{\alpha}_{\beta} + u^{\alpha}u_{\beta}$, we arrive at
\begin{equation}
(w + b^2)u_{\nu}\partial_i u^{\nu} + \partial_i\left(p + \frac{b^2}{2}\right) - b_{\nu}\partial_i b^{\nu},
\end{equation}
where $i = r, \theta$. We can rewrite this equation in terms of the specific angular momentum $l$ and the angular velocity $\Omega$, to obtain
\begin{equation}\label{eq:diff_ver}
\partial_i(\ln u_t|) - \frac{\Omega \partial_i l}{1-l\Omega} + \frac{\partial_i p}{w} + \frac{\partial_i(\mathcal{L}b^2)}{2\mathcal{L}w} = 0,
\end{equation}
where 
\begin{equation}
\mathcal{L} = g_{t\phi}^2 - g_{tt}g_{\phi\phi}.
\end{equation}
To integrate Eq.~\eqref{eq:diff_ver} we first assume a barotropic equation of state $w = w(p)$ of the form
\begin{equation}\label{eq:eos_fluid}
p = K w^{\kappa},
\end{equation}
with $K$ and $\kappa$ constants.

Then, we define the magnetic pressure as $p_{\mathrm{m}} = b^2/2$. Also, we can define $\tilde{p}_{\mathrm{m}} = \mathcal{L} p_{\mathrm{m}}$ and $\tilde{w} = \mathcal{L} w$, so we can write an analogue of \eqref{eq:eos_fluid} for $\tilde{p}_{\mathrm{m}}$
\begin{equation}\label{eq:eos_mag_tilde}
\tilde{p}_{\mathrm{m}} = K_{\mathrm{m}} \tilde{w}_{\mathrm{m}}^{\eta},
\end{equation}
or, in terms of the magnetic pressure $p_{\mathrm{m}}$
\begin{equation}\label{eq:eos_mag}
p_{\mathrm{m}} = K_{\mathrm{m}} \mathcal{L}^{\eta-1} w^{\eta},
\end{equation}
where $K_{\mathrm{m}}$ and $\eta$ are constants.

This particular choice of relationships $w = w(p)$ and $\tilde{w} = \tilde{w}(\tilde{p}_{\mathrm{m}})$, fulfill the general relativistic version for a toroidal magnetic field of the von Zeipel theorem \citep{vonZeipel:1924, Zanotti:2015}, that means the surfaces of $\Omega$ and $l$ constant coincide.

We can rewrite equation \eqref{eq:diff_ver} and integrate it to give
\begin{equation}\label{eq:pre_full_int}
\ln |u_t| - \int^l_0 \frac{\Omega \mathrm{d}l}{1 - \Omega l} + \int^p_0 \frac{\mathrm{d}p}{w} + \int^{\tilde{p}_{\mathrm{m}}} \frac{\mathrm{d}\tilde{p}_{\mathrm{m}}}{\tilde{w}} = \mathrm{constant}.
\end{equation}
On the surface of the disk, and particularly on its inner edge,
\begin{equation}
p = \tilde{p}_{\mathrm{m}} = 0, \;\;\; u_t = u_{t, \mathrm{in}}, \;\;\; l = l_{\mathrm{in}}
\end{equation}
then, the constant of integration is given by
\begin{equation}
\mathrm{constant.} = \ln |u_t| - \int^l_{\mathrm{in}} \frac{\Omega \mathrm{d}l}{1 - \Omega l}.
\end{equation}
Defining the total potential $W$ \citet{Abramowicz:1978}, we can write
\begin{equation}\label{eq:potential}
W - W_{\mathrm{in}} = \ln|u_t| - \ln|u_{t,\mathrm{in}}| - \int^{l}_{l_{\mathrm{in}}} \frac{\Omega \mathrm{d}l}{1 - \Omega l},
\end{equation}
where $W_{\mathrm{in}}$ is the potential on the inner edge of the disk. With this definition, we can write \eqref{eq:pre_full_int} as
\begin{equation}\label{eq:full_int}
W - W_{\mathrm{in}} = \int^p_0 \frac{\mathrm{d}p}{w} + \int^{\tilde{p}_{\mathrm{m}}} \frac{\mathrm{d}\tilde{p}_{\mathrm{m}}}{\tilde{w}},
\end{equation}
and
\begin{equation}\label{eq:pre_enthalpy_eq}
W - W_{\rm{in}} + \frac{\kappa}{\kappa - 1}\frac{p}{w} + \frac{\eta}{\eta - 1}\frac{p_{\mathrm{m}}}{w} = 0,
\end{equation}
and then, replacing $p$ and $p_{\mathrm{m}}$ by equations \eqref{eq:eos_fluid} and \eqref{eq:eos_mag},
\begin{equation}\label{eq:enthalpy_eq}
W - W_{\rm{in}} + \frac{\kappa}{\kappa - 1} K w^{\kappa - 1} + \frac{\eta}{\eta - 1}K_{\mathrm{m}}(\mathcal{L} w)^{\eta - 1} = 0,
\end{equation}
which relates the distribution of potential with the distribution of enthalpy.

>>>>

The investigation of such systems, either by analytical or numerical means, may rely on the ability to construct suitable 
representations based on physical assumptions. The construction of equilibrium models of stationary disks around black holes has a long tradition (see~\cite{Abramowicz:2013} and references therein). In particular, the so-called ``Polish doughnuts"~\citep{Abramowicz:1978,Kozlowski:1978} provide a  very general method to build equilibrium configurations of perfect fluid matter  
 orbiting around a Kerr black hole. This fully relativistic model assumes that the disk is non-magnetized and that the matter obeys a  constant specific angular momentum distribution. This method was extended by~\cite{Komissarov:2006}  by adding a purely azimuthal magnetic 
 field to build magnetized tori around rotating black holes. Assuming different distributions of angular momentum in the disks, \citet{Qian:2009} presented a method to build sequences of black hole thick accretion disks in dynamical equilibria, restricted to the purely hydrodynamical case. Building on this work, we discuss here the extension of the
models of~\citet{Qian:2009} to disks endowed with a purely toroidal magnetic field. For the particular case of constant angular momentum distributions our method shows good agreement with the results of~\citet{Komissarov:2006}.

@@

The equations of ideal general relativistic MHD are $\nabla_{\mu} T^{\mu\nu} = 0$, $\nabla_{\mu} \,^\ast F^{\mu\nu} = 0$, and 
$\nabla_{\mu} (\rho u^{\mu}) = 0$, 
%\begin{equation}\label{eq:e-m_cons}
%\nabla_{\mu} T^{\mu\nu} = 0,
%\end{equation}
%\begin{equation}\label{eq:cons_faraday}
%\nabla_{\mu} \ast F^{\mu\nu} = 0,
%\end{equation}
%\begin{equation}\label{eq:continuity}
%\nabla_{\mu} (\rho u^{\mu}) = 0,
%\end{equation}
where $\nabla_{\mu}$ is the covariant derivative and
\begin{equation}\label{eq:e-m_tensor}
T^{\mu\nu} = (w + b^2)u^{\mu}u^{\nu} + \left(p + \frac{1}{2}b^2\right)g^{\mu\nu} - b^{\mu}b^{\nu},
\end{equation}
is the energy-momentum tensor of a magnetized perfect fluid, $w$ and $p$ being the fluid enthalpy and pressure, respectively. 
Moreover,
%\begin{equation}
$^\ast F^{\mu\nu} = b^{\mu}u^{\nu} - b^{\nu}u^{\mu}$
%\end{equation}
is the (dual of the) Faraday tensor relative to an observer with four-velocity $u^{\mu}$, and $b^{\mu}$ is the magnetic field in that frame, with
$b^2=b^{\mu}b_{\mu}$. Assuming the magnetic field is purely azimuthal, i.e.~$b^r = b^{\theta} = 0$,
and taking into account that the flow is stationary and axisymmetric, the conservation of the current density and of the Faraday tensor follow. Contracting Eq.~\eqref{eq:e-m_tensor} with the projection tensor $h^{\alpha}_{\beta} = \delta^{\alpha}_{\beta} + u^{\alpha}u_{\beta}$, we arrive at
\begin{equation}
(w + b^2)u_{\nu}\partial_i u^{\nu} + \partial_i\left(p + \frac{b^2}{2}\right) - b_{\nu}\partial_i b^{\nu}=0\,,
\end{equation}
where $i = r, \theta$. Following~\cite{Komissarov:2006} we rewrite this equation in terms of the specific angular momentum $l$ and the angular velocity $\Omega$, to obtain
\begin{equation}\label{eq:diff_ver}
\partial_i(\ln u_t|) - \frac{\Omega \partial_i l}{1-l\Omega} + \frac{\partial_i p}{w} + \frac{\partial_i(\mathcal{L}b^2)}{2\mathcal{L}w} = 0\,,
\end{equation}
where $\mathcal{L} = g_{t\phi}^2 - g_{tt}g_{\phi\phi}$.
To integrate Eq.~\eqref{eq:diff_ver} we first assume a barotropic equation of state $w = w(p)$ of the form
\begin{equation}\label{eq:eos_fluid}
p = K w^{\kappa},
\end{equation}
with $K$ and $\kappa$ constants.
Then, we define the magnetic pressure as $p_{\mathrm{m}} = b^2/2$, and introduce the definitions $\tilde{p}_{\mathrm{m}} = \mathcal{L} p_{\mathrm{m}}$ and $\tilde{w} = \mathcal{L} w$, in order to write an analog equation to Eq.~\eqref{eq:eos_fluid} for $\tilde{p}_{\mathrm{m}}$~\citep{Komissarov:2006}
\begin{equation}\label{eq:eos_mag_tilde}
\tilde{p}_{\mathrm{m}} = K_{\mathrm{m}} \tilde{w}_{\mathrm{m}}^{\eta},
\end{equation}
or, in terms of the magnetic pressure $p_{\mathrm{m}}$
\begin{equation}\label{eq:eos_mag}
p_{\mathrm{m}} = K_{\mathrm{m}} \mathcal{L}^{\eta-1} w^{\eta},
\end{equation}
where $K_{\mathrm{m}}$ and $\eta$ are constants.
This particular choice of relationships, $w = w(p)$ and $\tilde{w} = \tilde{w}(\tilde{p}_{\mathrm{m}})$, fulfill the general relativistic version of the von Zeipel theorem for a toroidal magnetic field~\citep{vonZeipel:1924, Zanotti:2015}, i.e.~the surfaces of constant $\Omega$ and constant $l$ coincide.

Integrating Eq.~\eqref{eq:diff_ver} we obtain
\begin{equation}\label{eq:pre_full_int}
\ln |u_t| - \int^l_0 \frac{\Omega \mathrm{d}l}{1 - \Omega l} + \int^p_0 \frac{\mathrm{d}p}{w} + \int^{\tilde{p}_{\mathrm{m}}} \frac{\mathrm{d}\tilde{p}_{\mathrm{m}}}{\tilde{w}} = \mathrm{const}.
\end{equation}
On the surface of the disk, and particularly on its inner edge, the conditions
$p = \tilde{p}_{\mathrm{m}} = 0, \; u_t = u_{t, \mathrm{in}}, \; l = l_{\mathrm{in}}$
are satisfied and, therefore, the integration constant is given by
\begin{equation}
\mathrm{const.} = \ln |u_t| - \int^l_{l_\mathrm{in}} \frac{\Omega \mathrm{d}l}{1 - \Omega l}\,.
\end{equation}
We can also introduce the total (gravitational plus centrifugal) potential $W$~\citep{Abramowicz:1978} and write the integral form of the equation of motion as
\begin{equation}\label{eq:potential}
W - W_{\mathrm{in}} = \ln|u_t| - \ln|u_{t,\mathrm{in}}| - \int^{l}_{l_{\mathrm{in}}} \frac{\Omega \mathrm{d}l}{1 - \Omega l},
\end{equation}
where $W_{\mathrm{in}}$ is the potential on the inner edge of the disk (in the equatorial plane). With this definition, we can write \eqref{eq:pre_full_int} as
\begin{equation}\label{eq:full_int}
W - W_{\mathrm{in}} = \int^p_0 \frac{\mathrm{d}p}{w} + \int^{\tilde{p}_{\mathrm{m}}} \frac{\mathrm{d}\tilde{p}_{\mathrm{m}}}{\tilde{w}},
\end{equation}
and
\begin{equation}\label{eq:pre_enthalpy_eq}
W - W_{\rm{in}} + \frac{\kappa}{\kappa - 1}\frac{p}{w} + \frac{\eta}{\eta - 1}\frac{p_{\mathrm{m}}}{w} = 0,
\end{equation}
and then, replacing $p$ and $p_{\mathrm{m}}$ by equations \eqref{eq:eos_fluid} and \eqref{eq:eos_mag},
\begin{equation}\label{eq:enthalpy_eq}
W - W_{\rm{in}} + \frac{\kappa}{\kappa - 1} K w^{\kappa - 1} + \frac{\eta}{\eta - 1}K_{\mathrm{m}}(\mathcal{L} w)^{\eta - 1} = 0,
\end{equation}
which relates the distribution of potential with the distribution of enthalpy.
